---
title: "Pirat"
author: 
- name: Lucas Etourneau
- name: Samuel Wieczorek"
package: Pirat
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc_float: true
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Pirat-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography:
  - biblio_vignette.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


<!-- ```{r style, echo = FALSE, results = 'asis'} -->
<!-- BiocStyle::markdown() -->
<!-- ``` -->



# Installation

`r Biocpkg("Pirat")` is an `R` package available via the 
[Bioconductor](http://bioconductor.org) repository for packages. 
<!-- You can install the release version by using the following commands  -->
<!-- in your `R` session: -->

<!-- ```{r "install", eval = FALSE} -->
<!-- if (!requireNamespace("BiocManager", quietly = TRUE)) { -->
<!--     install.packages("BiocManager") -->
<!-- } -->
<!-- BiocManager::install("cfTools") -->
<!-- ``` -->

You can install the development version from [GitHub](https://github.com/) :

```{r install_dev, eval = FALSE}
BiocManager::install("prostarproteomics/Pirat")
```


When calling one of the imputation methods for the first time after package
installation, the underlying python module and conda environment will be 
automatically downloaded and installed without the need for user intervention.
This may take several minutes, but this is a one-time installation. 

# Standard imputation with Pirat

Pirat is a single imputation pipeline including a novel statistical approach 
dedicated to bottom-up proteomics data. All the technical details and the 
validation procedure of this method are available on the corresponding 
preprint @Etourneau2023 .
To demonstrate its usage, we first load Pirat package and a subset of 
Bouyssie2020 dataset in the environment.

```{r load_data, echo = TRUE}
library(Pirat)
library(utils)
data(subbouyssie)
```

Note that `subbouyssie` is a list that contains two elements:

1. `peptides_ab`, the matrix of peptide log-2-abundances, with samples in rows 
and peptides in columns.
2. `adj`, the adjacency matrix between peptides and proteins, containing either 
booleans or 0s and 1s (no preprocessing or simplification is needed for this 
matrix, Pirat will automatically build the PGs from it).

Slight imputation variations may occur for peptides belonging to very large 
PGs, because the latter are randomly split into several smaller PGs with fixed 
size to reduce computational costs. Although this variability is too small to 
affect imputation quality, we fix the seed in this tutorial such that the user 
can retrieve exactly the same imputed values when running the notebook again.

```{r setseed}
set.seed(12345)
```

One can then impute this dataset with the following line

```{r impute}
imp.res <- my_pipeline_llkimpute(subbouyssie)
```

The result `imp.res` is a list that contains:

1. `data.imputed`, the imputed log-2 abundance matrix.
2. `params`, a list containing parameters $\Gamma$ and hyperparameters 
$\alpha$ and $\beta$.

You can check the imputed values here...

```{r test1}
head(imp.res$data.imputed[ ,seq(5)])
```

...and the computed parameters here.

```{r params}
imp.res$params
```
These parameters were originally proposed in @Chen2014, 
however no methods were proposed to find then 
automatically and without heuristics.
Note that a positive value for $\gamma_1$ indicates that a random 
left-truncation mechanism is present in the dataset.

# Intra-PG correlation analysis

Pirat has a diagnosis tool that compares distributions of correlations at 
random and those from same peptide groups (PGs).
We use it here on the complete Ropers2021 dataset.

```{r correlations}
data(ropers)
plot_pep_correlations(ropers, titlename = "Ropers2021")
```


 We see here that the blue distribution has much more weights on high 
 correlations than the red one,
 indicating that PGs should clearly help for imputation.
 
# Pirat extensions
 
To handle singleton PGs, Pirat proposes three extensions, on top of the 
classical Pirat approach.
Note that the -T extension can be applied to up to an arbitrary PG size.
To illustrate our examples, we use a subset of Ropers2021 dataset.
 
# -2, the 2-peptide rule
 
The -2 extension simply consists in not imputing singleton PGs. It can be used as following.

```{r pipeline_llkimpute}
data(subropers)
imp.res = pipeline_llkimpute(subropers, extension = "2")
```

We can then check that singleton peptides are not imputed (yet some may be 
already fully observed).

```{r impute4}
mask.sing.pg = colSums(subropers$adj) == 1
mask.sing.pep = rowSums(subropers$adj[, mask.sing.pg]) >= 1
imp.res$data.imputed[, mask.sing.pep]
```

# -S, samples-wise correlations

Pirat can leverage sample-wise correlations to impute the singleton peptides 
as following:

```{r my_pipeline_llkimpute2}
imp.res = my_pipeline_llkimpute(subropers, extension = "S")
```

Here the singleton peptides are imputed after the rest of the dataset, using 
the sample-wise covariance matrix obtained.


```{r impute2}
mask.sing.pg = colSums(subropers$adj) == 1
mask.sing.pep = rowSums(subropers$adj[, mask.sing.pg]) >= 1
imp.res$data.imputed[, mask.sing.pep]
```

# -T, transcriptomic integration

The last extension consists in using correlations between peptides and 
gene/transcript expression obtained from a related transcriptomic analysis. 
To use this extension, the list of the dataset must contain:

* `rnas_ab`, an log2-normalized-count table of gene or transcript expression, 
for which samples are either paired or related (*i.e.*, from the same 
experimental/biological conditions).
* `adj_rna_pg`, a adjacency matrix between transcripts or genes and PGs, 
containing either booleans or 0s and 1s.

`ropers` proteomic and transcriptomic samples are paired (*i.e.* the same 
biological samples were used for each type of analysis). Thus Pirat-T can be 
used as following:

```{r my_pipeline_llkimpute3}
imp.res = my_pipeline_llkimpute(subropers,
    extension = "T",
    rna.cond.mask = seq(nrow(subropers$peptides_ab)),
    pep.cond.mask = seq(nrow(subropers$peptides_ab)),
    max.pg.size.pirat.t = 1)
```

Only a few peptides have been used to fit the prior variance distribution in 
$\Sigma$, as we use a small subset from the original Ropers2021 dataset for 
demonstration purpose. However, it is preferable
in general to use the full dataset for both hyperparameter and $\Gamma$ estimation. 

It gives the following imputed singletons:

```{r data.imputed3}
mask.sing.pg = colSums(subropers$adj) == 1
mask.sing.pep = rowSums(subropers$adj[, mask.sing.pg]) >= 1
imp.res$data.imputed[, mask.sing.pep]
```

On the other hand, if proteomic and transcriptomic samples are not paired but 
are derived from a same biological/experimental condition. Pirat-T can be used 
by adapting the mask related to samples in each type of analysis (here, both 
proteomic and transcriptomic datasets have 6 different conditions in the same 
order with 3 replicates each):

```{r my_pipeline_llkimpute_T}
imp.res = my_pipeline_llkimpute(subropers,
                             extension = "T",
                             rna.cond.mask = rep(seq(6), each = 3),
                             pep.cond.mask = rep(seq(6), each = 3),
                             max.pg.size.pirat.t = 1)
```

Also, it is possible to apply transcriptomic integration up to an arbitrary 
size of PG, simply by 
changing parameter `max.pg.size.pirat.t` in `my_pipeline_llkimpute()` to the 
desired limit PG size (*e.g.* `+Inf` for whole dataset).

# 5. Session info

```{r sessionInfo}
sessionInfo()
```

# References
