---
title: "Pirat"
author: "Lucas Etourneau, Samuel Wieczorek"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Pirat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Table of content

1. [Pirat installation](#dunno)

## Pirat installation

Pirat requires a particular Python environment to work. Follow instructions in the README file to install Pirat correctly.

## Standard imputation with Pirat

Load Pirat package and a reduced part of Bouyssie2020 dataset in the environment.
```{r, eval=F}
library(Pirat)
data("bouyssie")
```

Note that `bouyssie` is actually a list that contains two elements:

1. `peptides_ab`, the matrix of peptide log-2-abundances, with samples in rows and peptides in columns.
2. `adj`, the adjacency matrix between peptides and proteins, containing etiher booleans or 0s and 1s.

You can then impute this dataset with the following line
```{r, eval=F}
imp.res <- pipeline_llkimpute(bouyssie)
```

The result `imp.res` is a list that contains:

1. `data.imputed`, the imputed log-2 abundance matrix.
2. `params`, a list containing parameters $\Gamma$ and hyperparameters $\alpha$ and $\beta$.

Check the imputed dataset here...
```{r}
head(imp.res$data.imputed[1:5])
```

...and the computed parameters here.
```{r}
imp.res$params
```
Note that a positive value for $\gamma_1$ indicates that a random left-truncation mechanism
is present in the dataset.

## Intra-PG correlation analysis

Pirat has a diagnosis tool that compares distributions of correlations 




Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
